Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-06-14 pointbypoint.tex

%\documentclass[sn-basic]{sn-jnl}
\documentclass[10pt]{article}
\usepackage{natbib}
\usepackage[margin=1in]{geometry}
\bibliographystyle{plainnat}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{soul}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{wrapfig}

\newcommand{\comment}[1]{\textcolor{teal}{#1}}
\newcommand{\bruno}[1]{\textcolor{red}{#1}}

% TODO
%
% in pointbypoint, add description of why parallel tempering for lognormal and why not in gamma.
% additional description of prior elicitation
% future work in conclusion.
% Responses for already answered comments referencing previous comments.
%

@@

We note in passing that the moment conditions for the spectral distribution, say 
    $\Phi$, in the two-dimensional case, using the infinity norm, imply that 
    \[
        \int_0^{\pi/4} \Phi(d\theta) = \int_{\pi/4}^{\pi/2}
    \Phi(d\theta).
    \]
    The former looks like a fairly strong symmetry restriction that is unlike to 
    be realistic in practice. Thus, imposing moment conditions would probably 
    result in a very restrictive model.

\item \emph{Main contribution}: While I appreciate the interesting construction 
    based on the DPM of projected gammas (which itself could be of use for other 
    contexts beyond extremes?), one wonders about what would be the main added 
    value of it with respect to other existing Bayesian approaches 
    (\cite{boldi2007},\cite{guillotte2011},\cite{SaNa2014},\cite{hanson2017}).
    Keeping in mind the computational scope of Statistics and Computing, one 
    wonders if the main advantage is, for example, a computational one? This 
    needs to be clarified—especially in light of the previous comment on the 
    moment constraint. Indeed, one could use any of the previous approaches with 
    pseudo-angles based on the infinity norm.

\comment{Please review this response.  I've done a little googling, but I can't figure
    out how one would sample from the Bernstein polynomial model.  Maybe after a lot of
    integration you could sort out some inverse CDF transform?}

We thank the reviewer for inviting us to expound on differences between our approach
    and prior contributions. We have clarified the distinctions between our approach 
    and the prior contributions
    mentioned.  \cite{guillotte2011} is a strictly bivariate model which assumes
    unit Fr\'{e}chet margins.  \cite{boldi2007} similarly assumes unit Fr\'{e}chet
    margins, but builds the spectral density representation on the unit simplex with
    a finite mixture of Dirichlets.  \cite{SaNa2014} assumes the same, but implements
    a reversible jump MCMC to allow varying the number of mixture components.
    \cite{hanson2017} similarly assumes unit Fr\'{e}chet margins, and to represent
    the spectral density implements a mixture of Bernstein polynomial densities.
    The distinction as compared to these methods is three-fold: the projected gamma does
    offer computational advantages as inference is straightforward, and perhaps as 
    importantly it is simple to sample from.  As a Dirichlet process model, inference
    on component weights is also straightforward.  It is highly flexible, with no
    imposed restriction on the number of dimensions.  Finally, as stated in the previous
    comment, the moment restriction used in those papers is incompatible with our model.
    This allows us to consider a wider class of distributions on pseudo-angles.

    % Setting aside \cite{guillotte2011} which is a strictly bivariate
    % model, the named approaches consider batch extrema and by construction must
    % respect the moment constraint.  As noted, our approach is able to set that 
    % aside, allowing consideration of a wider class of distributions of pseudo-angles.
    % \bruno{\bf Where is this in the paper?}

@@

    % \comment{Hanson is clearly different.  They're using Bernstein polynomials 
    % effectively defined on $\mathbb{S}_1^{d-1}$.  We can demonstrate that 
    % projected gammas built on $\mathbb{S}_{p}$, $p \geq 10$ and $\mathbb{S}_1$ 
    % create \emph{very} different gamma shape parameters, for data created by 
    % the \emph{same} distribution.  I need to check again the other references, 
    % but based on their follow-on comment, I'm guessing the previous point needs 
    % to be addressed before we can properly justify this point.}

    % \bruno{Based on our conversation, you need to look carefully into each one 
    % of these references. it seems that most of them do not develop a PoT 
    % approach. But also, we use a kernel whose natural support is the positive 
    % cone (without using transformations) to build a mixture WRT to a random
    % measure. This provides the flexibility of mixture for a methods that has 
    % computational advantages and scales to moderately large problems, as 
    % illustrated in the paper.}

\item \emph{Numerical accuracy}:  Keeping in mind the computational scope of 
    Statistics and Computing one would hope for more numerical work than what is 
    presented in §5.1. As far as I understand what is presented are single run 
    experiments for each $d$? Monte Carlo evidence on the performance of the 
    procedure would be appreciated. On top of this, the current numerical exercise
    ignores the uncertainty that stems from step 1 of the procedure 
    (i.e. estimating the margins).

To address this concern, we have revised the simulation study to feature the 
    average rise in energy score over baseline, averaged over 10 replications.  
    The simulated datasets are generated as mixtures of multivariate lognormals, 
    projected onto the unit sphere.  As you say, this paper focuses on estimating
    the spectral distribution, and ignores uncertainty from estimating marginal
    parameters.  We have clarified as such.
    \comment{I'm not sure how to address this.  I added a comment in the paper
    that explicitly stated that uncertainty in marginal parameters is outside the
    scope of the paper.  As a practical concern, every paper I have seen assumed
    pre-standardized marginals, which implies estimating marginal parameters and
    ignoring the uncertainty stemming from that estimation.}

@@

\item \emph{Prior and MCMC}: A lot is left to the reader in §5 in terms of
    prior specification and MCMC settings.  At least some further details would 
    be required.

    \comment{We thank you for drawing our attention to this omission.  We have included the
    details of the prior parameters at the start of section 5.}

    % \comment{The form of the prior is specified.  I'm guessing he's referring to 
    % the actual values used for computation?  $\bm{\mu}_0 = \bm{0}$, 
    % $\nu = d + 50$, $\bm{\Psi} = \nu I_d$, and so on?}

    % \bruno{You need to provide more details here.}

@@

    \item About the standardization, Eq. (3): please introduce some background. In the paper it is
        introduced as \emph{’the standardization’} and it is unclear what is the relation between this 
        and the definition of the GPD (beginning of Section 2), unless one already knows about marginal
        standardization methods in multivariate EVT. In any case there are several ways to perform 
        marginal standardization in multivariate EVT.

        We have rewritten the material at the beginning of Section 3, explicitly mentioning the marginal
        univariate generalized Pareto, and how it is used to estimate the shape and scale parameters, 
        which are then used to standardize the observations.

    \item Section 3.2.1: Please provide some minimal background (or references) on the Dirichlet 
        Process Mixture, in particular about clusters of observations occurring naturally in this 
        model. This is indeed standard in Bayesian Non-parametrics, so there should be a precise 
        reference where it is explained.

        To ameliorate this shortfall, in addition to the foundational papers for the DP
        already included, we add \cite{muller2015} to provide a modern approachable reference on
        and associated clustering, as well as \cite{ascolani2022} which provides a discussion of
        robustness of clusters generated under the DP.

        % \bruno{Peter, you need to come up with additional, and more specific references here.}
        % \comment{We have \cite{Ferguson74}, \cite{Antoniak1974}, (foundational) \cite{neal2000},
        % (methodological relative to auxiliary Gibbs) and \cite{escobar1995} (methodological 
        % relative to concentration parameter sampling).  I'm not sure what additional we need? 
        % I added an additional reference \citep{ascolani2022}, which concerns the robustness of
        % clustering, but I'm not sure of the relevance.}

@@

        \item Another issue is that it is not clear in the present form of the paper, what is the
        added value of fitting this parametric model with the Bayesian framework and the Gibbs 
        sampler, if the output is mainly bivariate densities and extremal dependence coefficients 
        which could be obtained by standard non parametric methods with very low computational cost?

        \bruno{This is an unfair comment. We do use the full multivariate distributiont to come up
        with the results we present. I'll work on an answer to this one.}
    \end{enumerate}

\item \emph{Comparison with other methods}: The only comparison study displayed in the paper 
    takes place within the suggested model, only with different prior choice. This is not sufficient 
    for performance comparison in a methodological paper. One may e.g. think of comparing with the
    results obtained with the empirical angular measure or smoothed versions of it in terms of 
    extremal dependence and conditional survival probability.

    We have included for comparison some other models established on $\mathcal{S}_{1}^{d-1}$;
        notably the pairwise betas model developed in \cite{sabourin2013}, as well
        as the Dirichlet distribution.  We have investigated several R and python packages to
        compute empirical angular measure, or other means of estimating the spectral distribution.
        To be clear, our analysis requires the ability to generate a sample from a posterior
        predictive distribution, or at least sample from a fitted density, projected onto 
        $\mathbb{S}_{\infty}^{d-1}$. Most packages implement a bivariate angular density, 
        including \verb|ExtremalDep|~\citep{ExtremalDep}, \verb|BAMBI|~\citep{BAMBI}, 
        \comment{expand list}.  Some, such as
        \verb|evd|~\citep{evd} implement a multivariate generalized Pareto density with a 
        parametric dependence structure, but I didn't see a way to learn the dependence
        structure in isolation, as well as sample from it.  I found 
        \cite{BMAmevt} as an example of a true multivariate density on $\mathbb{S}_1^{d-1}$, 
        and on first attempt to use
        it I encountered a bug with its output (for distributions with more than 3 dimensions, 
        the dimension of the output parameter space became larger than it should be).  Upon 
        conferring with the package maintainer and author, they identified the source of the
        bug and fixed it.  For this reason, we are able to include it as a point of comparison.

@@

    \bruno{Well, here we have to bite the bullet. Hopefully there some software that allows us
    to fit the empirical angular measure.}

    \item \emph{Experiments}:
        \begin{enumerate}
            \item Computational Evidence missing: The convergence of the Gibbs sampler is not 
            discussed in the paper.

            We thank the reviewers for bringing this omission to our attention.  We have now
            included a discussion of MCMC convergence among the models in section 5.

            % \bruno{This one was pointed out by the other referee. So, we need to come up with something.}

            % \comment{\st{I'm not sure how to demonstrate convergence in the case of the DP.  I can
            % I can demonstrate convergence for certain quantities (e.g., the hierarchical mean,
            % and log-determinant of hierarchical covariance matrix, a given observation's distribution
            % of parameters, etc.), but in the general stick-breaking configuration, there is no concern
            % given to label switching.  The variational approach will probably find a label arrangement
            % and stick with it.}}

            % \comment{I'm working on adding log-density over time plots to demonstrate convergence.}

@@

            \item Figure 2: The scores from the ‘baseline’ should outperform the scores of
            the fitted models, because the baseline consists of samples generated from the 
            true distribution. This is not the case in Figure 2, which deserves an explanation 
            (it might be a too small sample size, or some coding error at worst).

            We have converted the graphs to use average rise in score over baseline (as calculated
            by average of model on 10 distributions), rather than the score directly on a single
            sample. We have verified that instances of displayed values dropping below zero are 
            within MCMC error of zero.\comment{I need to say this better.}

### 2023-06-14 conclusion.tex
pgpareto github
statistics and computing submission

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto described in \cite{ferreira2014}
  to establish a useful representation of its dependence structure, through
  the distribution of its angular component, which is supported on the positive orthant of the unit
  hypersphere under the $\mathcal{L}_{\infty}$ norm.  Due to the 
  inherent difficulty of obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$
  our method projects distributions, supported on the positive quadrant and
  based on products of independent gammas,  onto the manifold
  ${\mathbb S}_{p}^{d-1}$. Samples of the resulting probability distribution 
  are then projected onto ${\mathbb S}_{\infty}^{d-1}$.  
  As ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as $
  p\to\infty$, we expect the resampling to be efficient for large enough
  $p$. In fact, our exploration of the simulated and real data indicate
  that the procedure is robust to the choice of moderately large values
  of $p$.

We explored the behavior of the proposed model using simulated data from
    mixtures of projected Gammas with varying degrees of complexity---both in terms of dimensionality and 
    in number of mixture components.  From this, we learned that the additional flexibility 
    offered by varying the $\beta$ term---the rate parameter of the Gamma distribution---does not result 
    in increased model fidelity in a mixture setting.  Within the tested range, between 3 and 20 
    dimensions, we observed that the additional information provided by a log--normal 
    centering distribution did not translate to additional model fidelity.  However, for 
    real data, with 47 dimensions, we did observe an effect.

The computations in this paper were performed on a desktop computer with an AMD Ryzen 5950X processor.
  The program is largely single-threaded, so computation time is not dependent on available core count.  In each
  case, we run the MCMC chain for 50\,000 iterations, with a burn-in of 20\,000 samples.  Fitting the PG--G model
  on the ERA5 dataset took approximately 30 minutes.  Work is in progress to optimize the code, and explore
  parallelization where possible.  We are also exploring alternative computational approaches that will make it
  feasible to tackle very high dimensional problems. In fact, to elaborate on the study of IVT, there is a need 
  to consider several hundreds, if not thousands, of grid cells over the Pacific Ocean in order to obtain a 
  good description of atmospheric events responsible for large storm activity over California.

% {\bf ( describe
% the computer)} and in each case we run the MCMC for {\bf XXX} iterations
% with a burn in of {\bf XXX} samples. For the large simulated data and the
% ERA5 dataset the process of fitting the model took {\bd YYY} minutes. Work is in progress to optimize the
% code to perform the MCMC {\bf (insert some comment here if you think it is appropriate)}. We are also 
% exploring alternative computational approaches that will make it feasible to tackle very high dimensional
% problems. In fact, to elaborate on the study of IVT, there is a need to consider several hundreds, if not
% thousands, of grid cells over the Pacific Ocean in order to obtain a good decsription of atmospheric events responsible for large storm activity over California.

%We used integrated vapor transport data---a measure of water in the atmosphere---from the ERA-Interim and ERA5
%    models as a motivating example for our model.  This data subsets California into a series of grid cells,
%    and provides daily IVT values for each cell.  For the ERA-Interim data, having determined from
%    Table~\ref{tab:dev} that the optimal model presented was PRG--G, we used this fitted model to develop
%    $\Chi$, a measure of pairwise extremal dependence between grid cells.  This provides a useful summary
%    statistic for describing the relationship between two sites.  Granted this relationship is pairwise, it
%    can not represent the full dependence structure; so recognizing these limits, we also presented 
%    conditional survival 
%    functions---$\text{P}\left[Z_{\ell} > z_{\ell}\mid Z_{\neg\ell} > z_{\neg\ell}\right]$---which
%    provide a more complete understanding of the dependence structure of $\bm{Z}$.

>>>>

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto 
    described in \cite{ferreira2014} to establish a useful representation of 
    its dependence structure, through the distribution of its angular component,
    which is supported on the positive orthant of the unit hypersphere under 
    the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
    obtaining a distribution with support on ${\mathbb S}^{d-1}_\infty$ our 
    method projects distributions, supported on the positive quadrant and 
    based on products of independent gammas,  onto the manifold 
    ${\mathbb S}_{p}^{d-1}$. Samples of the resulting probability distribution 
    are then projected onto ${\mathbb S}_{\infty}^{d-1}$.  As 
    ${\mathbb S}_{p}^{d-1}$ converges to ${\mathbb S}_{\infty}^{d-1}$ as 
    $p\to\infty$, we expect the resampling to be efficient for large enough
    $p$. In fact, our exploration of the simulated and real data indicate
    that the procedure is robust to the choice of moderately large values
    of $p$.

We explored the behavior of the proposed model using simulated data from
    mixtures of projected Gammas with varying degrees of complexity---both 
    in terms of dimensionality and in number of mixture components.  From this,
    we observed that the additional flexibility offered by varying the $\beta$
    term---the rate parameter of the gamma distribution---can offer additional
    model fidelity.  This result is not borne out in our real data example,
    for which the simpler model is vastly preferred.  Additionally, we observe
    that the choice of the log-normal centering distribution and additional
    flexibility it offers did not translate to additional model fidelity.  This
    is borne out in the real data, where on the ERA--Interim dataset it performed
    only comparably, but on ERA--5 it performed significantly worse.

    % From this,
    % we learned that the additional flexibility offered by varying the $\beta$ 
    % term---the rate parameter of the Gamma distribution---does not result in 
    % increased model fidelity in a mixture setting.  Within the tested range, 
    % between 3 and 20 dimensions, we observed that the additional information 
    % provided by a log--normal centering distribution did not translate to 
    % additional model fidelity.  However, for real data, with 47 dimensions, 
    % we did observe an effect.

The computations in this paper were performed on a desktop computer with an 
    AMD Ryzen 5950X processor. The program is largely single-threaded, so 
    computation time is not dependent on available core count.  In each case, 
    we run the MCMC chain for 50\,000 iterations, with a burn-in of 40\,000 
    samples.  Fitting the PG--G model on the ERA5 dataset took approximately 
    15 minutes.  Work is in progress to optimize the code, and explore 
    parallelization where possible.  We are also exploring alternative 
    computational approaches that will make it feasible to tackle very high 
    dimensional problems. In fact, to elaborate on the study of IVT, there is 
    a need to consider several hundreds, if not thousands, of grid cells over
    the Pacific Ocean in order to obtain a good description of atmospheric 
    events responsible for large storm activity over California.  Moving beyond
    this model

% §6 calls for a view of the authors on what 
%     could be related exercises—or even extensions of this research and future 
%     work. Using the proposed Bayesian multivariate PoT construction for 
%     devising a regression model for an an extreme value response on an extreme 
%     value covariate is one example of a natural follow-up; see 
%     \cite{carvalho2022} for related ideas. Conditional versions of the angular 
%     measure (e.g. \cite{carvalho2016}, \cite{castro2018}, \cite{escobar2018}, 
%     \cite{mhalla2019}) would lead to a conditional multivariate 
%     peaks-over-threshold model, where dependence between exceedances could 
%     change along with a covariate. Also, perhaps the here proposed DPM of 
%     projected gammas—as well as predictor-dependent versions of it—could be of 
%     interest in other settings beyond angular measures?

% EOF

### 2023-06-14 evaluation.tex
OL 2023_10_15

In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider 
    the theory of proper scoring rules developed in \cite{gneiting2007}. As mentioned in 
    Section~\ref{subsec:projgamma}, our approach does not provide a density on 
    ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria 
    to sample-based approaches.  To this end, we employ the \emph{energy score} criterion introduced
    therein.
    % the We consider two scoring methods:  \emph{posterior predictive loss} (PPL) introduced in 
    % \cite{gelfand1998}, and \emph{energy score} (ES) introduced in \cite{gneiting2007}.

% \subsection{Posterior predictive loss}
% The posterior predictive loss criterion is introduced in \cite{gelfand1998}.  When we
%   assume a squared error loss function, then for the $i$th observation, the posterior predictive
%   loss criterion is computed as
%   \begin{equation}
%     \label{eq:ppl}
%     \begin{aligned}
%       &\hspace{-0.5cm}S_{\text{PPL}}^k\left(P,x_i\right) \\
%       &= \text{Var}_P\left[X_i\right] + \frac{k}{k+1}\left(\text{E}_P\left[X_{i}\right] - x_i\right)^2
%     \end{aligned}
%   \end{equation}
%   where $X_i$ is a random variable from the posterior predictive distribution for $x_i$.  The
%   scalar $k$ is a weighting factor that scales the importance of goodness of fit
%   relative to precision.  In fact a small $\text{Var}(X_{i})$ indicates high
%   precision, and a small $(\text{E}[X_{i}] - x_{i})^2$  indicates a good fit.  In our analysis,
%   we take the limit as $k\to\infty$, weighting both
%   equally. Overall, the smaller the score, the better.  Note that the PPLC score is defined 
%   for a univariate distribution.  As we are dealing
%   with a multivariate distribution, we will be taking the posterior predictive loss
%   summed over all dimensions.  That is,
%   \begin{equation}
%     \label{eq:ppl2}
%     \begin{aligned}
%       &\hspace{-0.5cm}S_{\text{PPL}}\left(P,\bm{x}_i\right) \\
%       &= \frac{1}{d}\sum_{\ell = 1}^d\left[\text{Var}_P\left[X_{i\ell}\right] + \left(\text{E}_P\left[X_{i\ell}\right] - x_{i\ell}\right)^2\right].
%     \end{aligned}
%   \end{equation}
%   We report the average $S_{\text{PPL}}$ taken from all observed data.

% \subsection{Energy score}
% We focus on \emph{energy scores} defined for a general probability distribution $P$, with finite expectation, as 
The energy score criterion defined for a general probability distribution $P$, with finite expectation, is developed
    as
    \begin{equation}
    \label{eq:es}
    \begin{aligned}
    &\hspace{-0.5cm}\text{S}_{\text{ES}}\left(P, \bm{x}_i\right) \\
    &=\text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
                                    \frac{1}{2}\text{E}_p \left[g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)\right],
    \end{aligned}
    \end{equation}
where $g$ is a kernel function. The score defined in Equation \eqref{eq:es} can be evaluated
  using samples from $P$, with the help of the law of large numbers.
  Moreover, Theorem~4 in \cite{gneiting2007}, states that if $g(\cdot,\cdot)$ is a negative 
  definite kernel, then $\text{S}(P,\bm{x})$ is a \emph{proper} scoring rule.  Recall that a 
  real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, 
  and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, and any 
  collection $a_1,\ldots,a_n\in{\mathbb R}$ such that  $\sum_{i = 1}^na_i = 0$.

In a Euclidean space, these conditions are satisfied by the Euclidean distance \citep{berg1984}. 
    However, for observations on different faces of ${\mathbb S}_{\infty}^{d-1}$, Euclidean 
    distance will under-estimate the geodesic distance, the actual distance required to travel 
    between the two points.  Let
\begin{equation*}
    {\mathbb C}_{\ell}^{d-1} = \lbrace \bm{x} : \bm{x} \in {\mathbb S}_{\infty}^{d-1}, x_{\ell} = 1\rbrace
\end{equation*}
comprise the $\ell$th \emph{face}.  For points on the same face, Euclidean distance 
  corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.  
  For points on different faces, the Euclidean distance is a lower bound for that length.

@@

To calculate the energy score we define the kernel 
    \[  
    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
    \]
    where $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and 
    $\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
%\end{prop}
%{\em Proof:}
%When $\bm{a}$ and $\bm{b}$ are on the same face, then $\ell = \jmath$, and $g$ is the Euclidean distance between $\bm{a}$ and $\bm{b}$, which is a  negative definite kernel.  When $\bm{a}$ and $\bm{b}$ are on separate
  %faces, let 
  %\[\bm{e} = 
    %\argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        %\pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
  %\]
  %We have that $g(\bm{a}, \bm{b}) = g(\bm{b}, \bm{a})$ due to the uniqueness of $\bm{e}$ and the symmetry of the Euclidean distance. Additionally, given $n$ and  a collection $\alpha_1,\ldots,\alpha_n\in{\mathbb R}$ such that  $\sum_{i = 1}^n\alpha_i = 0$, then
  %\begin{equation*}
    %\begin{aligned}
      %&\hspace{-0.5cm}\sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j g(\bm{x}_1,\bm{x}_2)\\
      %&= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j \bigg[\pnorm{\bm{e} - \bm{x}_1}{2} + \pnorm{\bm{x}_2 - \bm{e}}{2}\bigg]\\
      %&= \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{e} - \bm{x}_1}{2}\\
      %&\hspace{1cm}+ \sum_{i = 1}^n\sum_{j = 1}^n \alpha_i\alpha_j\pnorm{\bm{x}_2 - \bm{e}}{2}\\
      %&\leq 0
    %\end{aligned}
  %\end{equation*}
  %as the Euclidean norm is a negative definite kernel.$\hfill\Box$
Evaluating $g$ as described requires the solution of a $(d-2)$-dimensional 
  optimization problem.  The following proposition provides a straightforward solution.
\begin{prop}\label{prop:rot}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, 
        for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the 
  $\ell$th axis produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for 
  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$ and let
\[\bm{e} = 
    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
\]
We then have that
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} 
        - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
    \end{aligned}
  \end{equation*}
  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the 
  same hyperplane. $\hfill\Box$

@@

Proposition \ref{prop:rot} provides a computational efficient way to evaluate the proper
    scoring rule $S_{\text{ES}}$ defined on ${\mathbb S}_\infty^{d-1}$, for each 
    observation. For the purpose of model assessment and comparison, we report the 
    average $S_{\text{ES}}$ taken across all observed data.  In this form, the 
    interpretation of $S_{\text{ES}}$ is similar to that of $S_{\text{PPL}}$: The smaller 
    the score, the better.

>>>>

In order to assess and compare the estimation of a distribution on ${\mathbb S}_\infty^{d-1}$ we consider 
    the theory of proper scoring rules developed in \cite{gneiting2007}. As mentioned in 
    Section~\ref{subsec:projgamma}, our approach does not provide a density on 
    ${\mathbb S}_{\infty}^{d-1}$, restricting our ability to construct model selection criteria 
    to sample-based approaches.  To this end, we employ the \emph{energy score} criterion introduced
    therein.

The energy score criterion defined for a general probability distribution $P$, with finite expectation, 
    is developed as
    \begin{equation}
    \label{eq:es}
    \begin{aligned}
    &\hspace{-0.5cm}\text{S}_{\text{ES}}\left(P, \bm{x}_i\right) \\
    &=\text{E}_p \left[g\left(\bm{X}_i, \bm{x}_i\right)\right] -
                                    \frac{1}{2}\text{E}_p \left[g\left(\bm{X}_i,\bm{X}_i^{\prime}\right)\right],
    \end{aligned}
    \end{equation}
where $g$ is a kernel function. The score defined in Equation \eqref{eq:es} can be evaluated
  using samples from $P$, with the help of the law of large numbers.
  Moreover, Theorem~4 in \cite{gneiting2007}, states that if $g(\cdot,\cdot)$ is a negative 
  definite kernel, then $\text{S}(P,\bm{x})$ is a \emph{proper} scoring rule.  Recall that a 
  real valued function $g$ is a negative definite kernel if it is symmetric in its arguments, 
  and $\sum_{i=1}^n\sum_{j=1}^na_ia_jg(x_i,x_j)\leq 0$ for all positive integers $n$, and any 
  collection $a_1,\ldots,a_n\in{\mathbb R}$ such that  $\sum_{i = 1}^na_i = 0$.

In a Euclidean space, these conditions are satisfied by the Euclidean distance \citep{berg1984}. 
    However, for observations on different faces of ${\mathbb S}_{\infty}^{d-1}$, the Euclidean 
    distance will under-estimate the geodesic distance, the actual distance required to travel 
    between the two points.  Let
\begin{equation*}
    {\mathbb C}_{\ell}^{d-1} = \lbrace \bm{x} : \bm{x} \in {\mathbb S}_{\infty}^{d-1}, x_{\ell} = 1\rbrace
\end{equation*}
    comprise the $\ell$th \emph{face}.  For points on the same face, the Euclidean distance 
    corresponds to the length of the shortest possible path in ${\mathbb S}_{\infty}^{d-1}$.  
    For points on different faces, the Euclidean distance is a lower bound for that length.

@@

To calculate the energy score we define the kernel 
    \[  
    g(\bm{a},\bm{b}) = \min_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
    \]
    where $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and 
    $\bm{b} \in {\mathbb   C}_{\jmath}^{d-1}$, for $\ell,\jmath\in \{1,\ldots, d\}$.
Evaluating $g$ as described requires the solution of a $(d-2)$-dimensional 
  optimization problem.  The following proposition provides a straightforward approach.
\begin{prop}\label{prop:rot}
    Let $\bm{a} \in {\mathbb C}_{\ell}^{d-1}$, and $\bm{b} \in {\mathbb C}_{\jmath}^{d-1}$, 
        for $\ell, \jmath \in \{1, \ldots , d\}$.  For $\ell\neq \jmath$, the 
  transformation $P_{\jmath\ell}(\cdot)$ required to rotate the $\jmath$th face along the 
  $\ell$th axis produces the vector $\bm{b}^\prime$, where
  \begin{equation}
    \label{eqn:rotation}
     \bm{b}^{\prime}_i = P_{\jmath\ell}(\bm{b}) = 
    \begin{cases}
        b_{i} &\text{for }i\neq \jmath,\ell\\
        1 &\text{for }i = \ell\\
        2 - b_{\ell} &\text{for }i = \jmath
    \end{cases}.
  \end{equation}
  Then $g(\bm{a},\bm{b}) = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}$.
\end{prop}
{\em Proof:}
Notice that for 
  $\bm{c} \in {\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}$,
     $\pnorm{\bm{b} - \bm{c}}{2} =  \pnorm{\bm{b}^{\prime} - \bm{c}}{2}$ and let
\[\bm{e} = 
    \argmin_{\bm{c} \in{\mathbb C}_{\jmath}^{d-1}\cap{\mathbb C}_{\ell}^{d-1}}\left\{ 
        \pnorm{\bm{c}-\bm{a}}{2} + \pnorm{\bm{b}-\bm{c}}{2} \right\}.
\]
We then have that
  \begin{equation*}
    \begin{aligned}
    g(\bm{a}, \bm{b}) &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b} - \bm{e}}{2}\\
    &= \pnorm{\bm{a} - \bm{e}}{2} + \pnorm{\bm{b}^{\prime} 
        - \bm{e}}{2} = \pnorm{\bm{a} - \bm{b}^{\prime}}{2}.
    \end{aligned}
  \end{equation*}
  The last equality is due to the fact that $\bm{a}$ and $\bm{b}'$ belong to the 
  same hyperplane. $\hfill\Box$

@@

Proposition \ref{prop:rot} provides a computational efficient way to evaluate the proper
    scoring rule $S_{\text{ES}}$ defined on ${\mathbb S}_\infty^{d-1}$, for each 
    observation. For the purpose of model assessment and comparison, we report the 
    average $S_{\text{ES}}$ taken across all observed data, and notice that the smaller 
    the score, the better.
